\chapter{Introduction}
\section{\sm}
The \sm is the fundamental theory of nature. There are three interactions included
\begin{itemize}
   \item electromagnetic
   \item weak
   \item strong
   \item Higgs boson exchange
\end{itemize}
Electromagnetic and weak interactions can be unified into electroweak interactions. 

In the Sun, all these three interactions and gravity are present
\begin{itemize}
   \item Photons reaching us clearly indicating QED's presence.
   \item Neutrinos produced in weak interaction. Four protons to two protons and two neutrons (Helium). 
      Only weak interaction can change the colour of quarks. 
   \item Binding of Helium via strong interaction and binding energy released as kinetic energy.
   \item Gravity brings protons together and at high temperature to give Helium.
\end{itemize}

\paragraph{Gauge theories} (Lie groups algebras)
\begin{itemize}
   \item EM: $\UEM$, $\UY$
   \item weak: $\SUL$
   \item strong: $\SUC$
\end{itemize}

Forces in quantum theories involve exchange particles spin $1$ vector bosons
\begin{itemize}
   \item Electromagnetism with photon $\gamma$
   \item Weak interaction with $W^{\pm}$, $W^0$ and $Z^0$ (mixture of $W^0$ and hyper charge) (discovered at CERN)
   \item Strong interaction with $g^a, a=1,\dots,8$ gluons (discovered at DESY)
\end{itemize}

\paragraph{particles with spin $ \frac{1}{2} $} \hspace{0pt} \\
Leptons
\begin{align*}
   \begin{pmatrix} \nu_e \\ e^- \end{pmatrix}_L , e^-_R ;\quad
   \begin{pmatrix} \nu_\mu \\ \mu^- \end{pmatrix}_L, \mu_R^-; \quad
   \begin{pmatrix} \nu_\tau \\ \tau^- \end{pmatrix}_L, \tau_R^- 
\end{align*}
Quarks
\begin{align*}
   \begin{pmatrix} u \\ d \end{pmatrix}_L, u_R, d_R; \quad
   \begin{pmatrix} c \\ s\end{pmatrix}_L, c_R, s_R; \quad
   \begin{pmatrix} t \\ b\end{pmatrix}_L, t_R, b_R
\end{align*}

One complete generation
\begin{align*}
   \begin{pmatrix} \nu_e \\ e^- \end{pmatrix}_L , e^-_R ;\quad 
   \begin{pmatrix} u \\ d \end{pmatrix}_L, u_R, d_R; \quad
\end{align*}
To remove any one part, then gauge theory is inconsistent. It is known as "anomaly". 

\paragraph{Higgs boson} $h^0$, spin $0$ \\
In the \sm, Higgs bosons are described by complex scalar fields $\begin{pmatrix} H^+ \\ H^0 \end{pmatrix} $ . $h^0$ is the only fundamental scalar in nature, as far as we know.

\section{Energy scales}
\begin{itemize}
   \item Binding energy of atoms $1-10 \si{\eV}$
   \item Binding energy of nucleons $\approx 1 \si{\mega\eV}$
   \item No known binding energy in particle physics
   \item Protons and neutrons $\approx \Lambda_{QCD} \approx \mathcal{O} (100 \si{\mega\eV})$
\end{itemize}

Particles have masses
\begin{itemize}
   \item Electron $m_e = \SI{511}{\kilo\eV}$
   \item Muon $m_\mu = \SI{105}{\mega\eV}$
   \item Tau $m_\tau = \SI{1.7}{\giga\eV}$
   \item Neutrinos $m_\nu < \SI{1}{\eV}$ 
   \item Quarks\footnote{mass of proton mainly comes from dynamical effect "gluon"} 
      \begin{itemize}
         \item $m_u \approx \SI{3}{\mega\eV}$
         \item $m_d \approx \SI{5}{\mega\eV}$ 
         \item $\dots$
      \end{itemize}
   \item Photon $m_\gamma = 0$
   \item Gluon $m_g = 0$
   \item Higgs $m_{Higgs} \approx \SI{125}{\giga\eV}$
\end{itemize}

\paragraph{Colliders}
\begin{itemize}
   \item Large Electron-Positron Collider (LEP) operating 1989-2000: $\SI{91}{\giga\eV} - \SI{200}{\giga\eV}$
   \item Tevatron($p\bar{p}$-collider) operating 1983-2011: $\SI{800}{\giga\eV} - \SI{2}{\tera\eV}$
   \item LHC operating since 2010: $\SI{7}{\tera\eV} - \SI{13}{\tera\eV}$
\end{itemize}

\section{Natural units}
It is convenient to use the unit system, where we define
\begin{align}
   \hbar = c = 1, \\
   k_B = 1.
\end{align}
Everything is expressed in term of powers of energy.
\begin{align*}
   \SI{1}{\femto \m} = \SI{1e-15}{\m} = \SI{5}{\per\giga\eV}   
\end{align*}
